Baseline Drift Estimation for Air Quality Data Using Quantile Trend Filtering

### 2019-11-20 Manuscript/response/response.tex

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath,amsfonts,amssymb,graphicx}
\usepackage[mathscr]{eucal}
\usepackage{stmaryrd}
\usepackage{amsbsy}
\usepackage{bm}
\usepackage[usenames]{color}
\usepackage[dvipsnames]{xcolor}
\usepackage{parskip} 
\usepackage{natbib}
\usepackage{comment}
%% For \subfloat command
\usepackage{subfig}

\newcommand{\fixme}[1]{{\color{red}\textbf{#1}}}
\newcommand{\T}[1]{\boldsymbol{\mathscr{\MakeUppercase{#1}}}} %tensor

%% Shorthand for Real
\newcommand{\Real}{\mathbb{R}}

\newcommand{\tr}{\operatorname{tr}} % Trace
\newcommand{\Kron}{\otimes} %Kronecker

\newcommand{\E}{\mathrm{E}}
\newcommand{\Cov}{\mathrm{Cov}}
\newcommand{\Corr}{\mathrm{Corr}}
\newcommand{\Ehat}{\hat{\E}}
\newcommand{\Covhat}{\hat{\Cov}}
\newcommand{\vect}{\mathrm{vec}}
\newcommand{\vech}{\mathrm{vech}}
\newcommand{\indep}{\;\, \rule[0em]{.03em}{.6em} \hspace{-.25em}
\rule[0em]{.65em}{.03em} \hspace{-.25em}
\rule[0em]{.03em}{.6em}\;\,}
\newcommand{\rank}{\mathrm{rank}}
\newcommand{\real}[1]{\Real \mathit{^{#1}}}
\newcommand{\trans}{^{\mbox{\tiny {\sf T}}}}
%\newcommand{\itrans}{^{\mbox{-\tiny {\sf T}}}}
\newcommand{\spn}{\mbox{span}}
\newcommand{\spc}{{\mathcal S}}
\newcommand{\diag}{\mathrm{diag}}
\newcommand{\cs}{\spc_{Y|\Xbf}}
\newcommand{\Var}{\mathrm{Var}}

%% Vectors
\newcommand{\V}[1]{{\bm{\mathbf{\MakeLowercase{#1}}}}} % vector
\newcommand{\VE}[2]{\MakeLowercase{#1}_{#2}} % vector element
\newcommand{\Vbar}[1]{{\bm{\M{A}r \mathbf{\MakeLowercase{#1}}}}} % vector
\newcommand{\Vhat}[1]{{\bm{\hat \mathbf{\MakeLowercase{#1}}}}} % vector
\newcommand{\Vtilde}[1]{{\bm{\tilde \mathbf{\MakeLowercase{#1}}}}} % vector
\newcommand{\Vn}[2]{\V{#1}^{(#2)}} % n-th vector
\newcommand{\VnE}[3]{{#1}^{(#2)}_{#3}} % n-th vector
\newcommand{\VtildeE}[2]{\tilde{\MakeLowercase{#1}}_{#2}} % vector element

%% Matrices
\newcommand{\Tra}{^{\sf T}} % Transpose
\newcommand{\M}[1]{{\bm{\mathbf{\MakeUppercase{#1}}}}} % matrix
\newcommand{\ME}[2]{\MakeLowercase{#1}_{#2}} % matrix element
\newcommand{\MC}[2]{\V{#1}_{#2}}
\newcommand{\Mr}[2]{\V{#1}_{#2}} % matrix row
\newcommand{\Mhat}[1]{{\bm{\hat \mathbf{\MakeUppercase{#1}}}}} % matrix
\newcommand{\Mtilde}[1]{{\bm{\tilde \mathbf{\MakeUppercase{#1}}}}} % matrix
\newcommand{\MhatC}[2]{\Vhat{#1}_{#2}} % matrix column
\newcommand{\Mbar}[1]{{\bm{\bar \mathbf{\MakeUppercase{#1}}}}} % matrix
\newcommand{\MbarC}[2]{\Vbar{#1}_{#2}} % matrix column
\newcommand{\Mn}[2]{\M{#1}^{(#2)}} % n-th matrix
\newcommand{\Mbarn}[2]{\Mbar{#1}^{(#2)}} % n-th matrix
\newcommand{\Mtilden}[2]{\Mtilde{#1}^{(#2)}} % n-th matrix
\newcommand{\MnTra}[2]{\M{#1}^{(#2){{\sf T}}}} % n-th matrix transpose
\newcommand{\MnE}[3]{\MakeLowercase{#1}^{(#2)}_{#3}} % n-th matrix element
\newcommand{\MnC}[3]{\V{#1}^{(#2)}_{#3}} % n-th matrix column
\newcommand{\MbarnC}[3]{\Vbar{#1}^{(#2)}_{#3}} % n-th matrix column
\newcommand{\MnCTra}[3]{\V{#1}^{(#2){{\sf T}}}_{#3}} % n-th matrix column transpose
\newcommand{\MCTra}[2]{\V{#1}^{{\sf T}}_{#2}} % n-th matrix column transpose
\newcommand{\prox}{\mathop{\rm prox}\nolimits}
\newcommand{\Pcal}{{\mathcal{P}}}
\newcommand{\greekbold}[1]{\mbox{\boldmath $#1$}}
\newcommand{\Sigmabf}{\greekbold{\Sigma}}

%% For comments to each other
\newcommand{\Remark}[2]{{\color{red}Remark from #1: #2}}
\newcommand{\Note}[1]{{{\color{red}{#1}}}}

\newcommand{\Eric}[2]{{\bf {\color{MidnightBlue}@Eric from #1: #2}}\xspace}
\newcommand{\Joe}[2]{{\bf {\color{Plum}@Joe from #1: #2}}\xspace}
\newcommand{\Halley}[2]{{\bf {\color{OliveGreen}@Halley from #1: #2}}\xspace}
\newcommand{\All}[2]{{\bf {\color{Sepia}@All from #1: #2}}\xspace}

\begin{document}
\title{Baseline Drift Estimation for Air Quality Data Using Quantile Trend Filtering}
\author{}
\date{}
\maketitle

\section*{General Remarks}

We thank the referees for their useful feedback and believe the manuscript is improved as a result.

In our revision we have denoted in \color{red}red \color{black} substantial newly added text. For the sake of readability we did not highlight minor edits in the revision.
Finally, we give point-by-point responses to referee comments below.
%\section*{Editor}
%
%\begin{quote}

%\Note{Should we mention here that we address reviewer 2's comments in the file?}
%
%\end{quote}

%\section*{Associate Editor}

\section*{Reviewer 1}

\begin{comment}
Authors present a method for quantile trend filtering with a non-crossing constraint. 
The use the ADMM algorithm to reduce the computing time. 
They compare their method with other approaches by numerical experiments. 
Finally, their approach is used on real data.

I really enjoy the paper. The method is interesting and well introduced. The approach requires different tuning parameters and I think that the choice of these parameters should be more discussed (see my comments below).

Major comments:
- The approach implies different tuning parameters: the penalty $\lambda$, the number of windows $W$, the bounds of the windows (l_w, u_w) and the degree of the polynomials (k). The selection of $\lambda$ is well described in Section 3. However, how the number of windows and the bounds of the windows are selected? In the paper, the degree of polynomials is k=2. Can the information criteria introduced in Section 3 be used to estimate k?

- In Section 3.1, authors said that the method can easily handle missing data. What are the assumption on the process of randomness (MAR, MCAR,...)?

- In Section 3.2, the BIC is introduced. I'm agree that for Bayesian quantile regression, the asymmetric Laplace likelihood is generally used. Does the choice of $\sigma$ impact the results? Moreover, in a Bayesian framework, a penalized regression can often be defined with a specific prior distribution on the parameters (L1-type regression can be defined by using a Laplace distribution as prior). Is it possible to define such prior for the problem defined by Equation (10)?

Minor comments:

- At the end of Section 3.2, maybe more details should be given about the properties of the eBIC for the large parameter spaces (is it more efficient than the BIC)?

- Figure 7,8 and 10 are not really clear. May be they can be replaced by tables.

\end{comment}

1. The approach implies different tuning parameters: the penalty $\lambda$, the number of windows $W$, the bounds of the windows $(l_w, u_w)$ and the degree of the polynomials $(k)$. The selection of $\lambda$ is well described in Section 3. However, how the number of windows and the bounds of the windows are selected?

\begin{quote}
The referee raises a good question. Our motivation for casting the problem as a linked set of estimations over overlapping windows is computational. Ideally there would be just one window to estimate the trend over, rather than $W > 1$ but as the trends in the timing experiments in Figure~5 indicate, using a single window will become computationally impractical if one wishes to fit a model over a very large set of time points. Our numerical experiments show a little less than a linear speed up in the number of windows used. Thus, if one wished to cut the overall running time to fit a family quantile trends by a roughly factor of 10, one would choose in the ballpark of 11 or 12 windows. The bounds of the windows and amount of overlap between windows depends on the application and should be larger than the expected time duration of transient events. We have added additional sentences addressing these points in the revised manuscript at the end of Section 2.
\end{quote}

2. In the paper, the degree of polynomials is $k=2$. Can the information criteria introduced in Section 3 be used to estimate $k$?

\begin{quote}
The expected degrees of freedom in the standard trend-filtering problem is given by the number of knots in $\V{\theta}$ plus $k+1$ \citep{Tib2014}. Consequently, a natural strategy to estimate $k$ is to use the eBIC criterion introduced in Section 3 but with $\nu$ equal to the number of knots in $\V{\theta}$ plus $k+1$. We have added additional sentences addressing these points in the revised manuscript at the end of Section 3.
\end{quote}

3. In Section 3.1, authors said that the method can easily handle missing data. What are the assumption on the process of randomness (MAR, MCAR,...)?

\begin{quote}
This is a good question. It turns out the penalized negative log-likelihood in Section 3.1 corresponds to making a MCAR assumption. We've added a short discussion of this in Section 3.1.
\end{quote}

4. In Section 3.2, the BIC is introduced. I'm agree that for Bayesian quantile regression, the asymmetric Laplace likelihood is generally used. Does the choice of $\sigma$ impact the results?

\begin{quote}
We did not find that the choice of $\sigma$ affected the results. The following figure plots the eBIC criteria against $\sigma$ for an example dataset. The value of $\lambda$ chosen is the same for all values of $\sigma$ compared, but the $\sigma$ used in the analysis (0.1) results in a eBIC value that rises more rapidly after achieving the minimum.

\begin{figure}[h!]
\centering
\includegraphics[width = 0.7\linewidth]{BIC_by_lambda_and_sigma.png}
\end{figure}
\end{quote}

5. Moreover, in a Bayesian framework, a penalized regression can often be defined with a specific prior distribution on the parameters (L1-type regression can be defined by using a Laplace distribution as prior). Is it possible to define such prior for the problem defined by Equation (10)?

\begin{quote}
Yes it is possible to define such a prior for the probelm defined by Equation (10). \cite{faulkner2018locally} proposed a few fully Bayesian variations of the trend filter utilizing shrinkage priors on the $k$th-order differences in values of the unknown target function. They used a Gaussian likelihood for the data corresponding to the squared loss used by \cite{Kim2009}. A fully Bayesian variation of the quantile trend filter could be implemented by substituting the Laplace likelihood for the Gaussian likelihood. %We have added these points to the discussion of prior work. 
We have added these points in the revised discussion.
\end{quote}

\subsection*{Minor comments:}

6. At the end of Section 3.2, maybe more details should be given about the properties of the eBIC for the large parameter spaces (is it more efficient than the BIC)?

\begin{quote}
Thank you for the suggestion. The eBIC has been proven to be model selection consistent in cases where the BIC can be model selection inconsistent. This difference between the eBIC and BIC manifests itself practically in the trend filtering problem in that the BIC selected models tend to be undersmoothed in comparison to the eBIC selected models. We have added sentence highlighting these advantageous properties of the eBIC for large parameter spaces in Section 3.2.

%``A limitation of the BIC, however, is that it may be inconsistent in the high-dimensional setting. This issue arises as a consequence of an implicit assumption in deriving the BIC that all models under consideration are equally likely. Consequently, in the context of variable selection methods such as the lasso, models that include more covariates (up to half of all possible covariates) will have a larger prior probability of being selected under the BIC criterion. In the context of this work, using the BIC may have a tendency to favor the selection of undersmoothed signals, as signals $\V{\theta}$ with more non-zero entries in the vector  $\Mn{D}{k+1}\V{\theta}$ will be assigned higher prior probabilities under the BIC.

%To address this issue, \cite{chen2008} modified the prior probabilities to dampen the prior weight on larger models, or in the context of this work, undersmoothed signals, and proposed 
%... They prove that the eBIC is model selection consistent under mild regularity conditions."
\end{quote}

7. Figure 7,8 and 10 are not really clear. May be they can be replaced by tables.

We have added tables with the results to the supplemental materials but feel that they are too long to include in the manuscript itself.

\section*{Reviewer 2}

\begin{comment}
The paper address the issue of estimating multiple quantile trends simultaneously in time series data. This problem is common in several fields from economics to medicine. The authors focused on the application to low cost air quality sensor. Thus the paper fits with AOAS policy. However, to improve the suitability, I suggest the authors to add more on how to employ their model and algorithm to data beyond to the application presented and which specific question this algorithm can address in different disciplines. While the problem to be addressed is clearly explained, the introduction miss a broader presentation of the topic to attract Statistician readers from differ fields. The authors claim that demixing problems is common to different applications with some references, thus they should comment more on this and reinforce the interest of the paper. Part of the technical material on the ADMM algorithm should be moved to appendices or online supplements.

The paper is correct. The algorithm is tested on simulated and real data supporting the validity of the results. The methodology and the results are well supported by figures and comments all over the test. Authors should check some typos spread in the manuscript.

Overall, the paper is sounds and interesting. It will benefit the community of Statistician working on low cost sensor data, an extremely increasing community. It will be of inspiration for scientists working with time series.
\end{comment}

The paper address the issue of estimating multiple quantile trends simultaneously in time series data. This problem is common in several fields from economics to medicine. The authors focused on the application to low cost air quality sensor. Thus the paper fits with AOAS policy.

8. However, to improve the suitability, I suggest the authors to add more on how to employ their model and algorithm to data beyond to the application presented and which specific question this algorithm can address in different disciplines. While the problem to be addressed is clearly explained, the introduction miss a broader presentation of the topic to attract Statistician readers from differ fields. The authors claim that demixing problems is common to different applications with some references, thus they should comment more on this and reinforce the interest of the paper.

\begin{quote}
We have expanded our discussion of baseline trend filtering applications to include electrocardiogram data, electrooculographic data, mass spectrometry, chromatography, macroeconomics, environmental science, laser-induced breakdown spectroscopy, and astronomy. We have also included one or more references for each example that requires baseline removal. These additions have been made on page 2 of the revision. We have also added a short paragraph at the end of the Section 6 on how our proposed method could potentially be used more generally for robust smoothing of time series data.
%\All{Eric}{Joe and Eric take lead in addressing, but please add any thoughts, Halley.} More applications: mass spec \cite{Du2006}. We don't explore it in this paper but we could use the framework for performing robust trend filtering by focusing on the quantiles close to the median. This might be useful for separating financial time series data into smooth trends and removing transient ups and downs. Maybe add this last observation to the discussion.
\end{quote}

9. Part of the technical material on the ADMM algorithm should be moved to appendices or online supplements.

\begin{quote}
We have moved part of the technical details of the ADMM algorithm to the supplementary materials.
\end{quote}

10. Authors should check some typos spread in the manuscript.

\begin{quote}
Thank you. We have checked and corrected for typos.
\end{quote}

\bibliographystyle{plainnat}
\bibliography{detrendify}
\end{document}

### 2019-11-20 Manuscript/response/main.tex
Submitted revision

% use option [preprint] to remove info line at bottom
% journal options: aop,aap,aos,aoas,ssy
% natbib option: authoryear
\documentclass[aoas]{imsart}
\RequirePackage{amsthm,amsmath,amsbsy,amsfonts}
\usepackage{bbm}
\usepackage{bm,graphicx,psfrag,epsf}
\usepackage{natbib}
\usepackage{url} % not crucial - just used below for the URL
\usepackage{placeins}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{subfig} %% For \subfloat command
\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}
% provide arXiv number if available:
%\arxiv{arXiv:0000.0000}

@@

\section{Introduction}
\label{sec:intro}
In the last decade, low cost and portable air quality sensors have enjoyed dramatically increased usage. These sensors can provide an un-calibrated measure of a variety of pollutants in near real time, but deriving meaningful information from sensor data remains a challenge \citep{snyder2013changing}. The ``SPod" is a low-cost sensor currently being investigated by researchers at the U.S. Environmental Protection Agency to detect volatile organic compound (VOC) emissions from industrial facilities \citep{thoma2016south}. Due to changes in temperature and relative humidity the output signal exhibits a slowly varying baseline drift on the order of minutes to hours. Figure \ref{fig:raw_spod} provides an example of measurements from three SPod sensors co-located at the border of an industrial facility. All of the sensors respond to the pollutant signal, which is illustrated by the three sharp transient spikes at 11:32, 14:10, and 16:03. However, the baseline drift varies from one sensor to another, obscuring the detection of the peaks that alert the intrusion of pollutants. We show later that by estimating the baseline drift in each sensor and  removing it from the observed signals, peaks can be reliably detected from concordant residual signals from a collection of SPods using a simple data-driven thresholding strategy. Thus, accurately demixing a noisy observed time series into a slowly varying component and a transient component can lead to greatly improved and simplified downstream analysis.

@@

While this work is motivated by the analysis of data from low cost air quality sensors, the problem of demixing nosy time series into trends and transients is ubiquitous across many fields of study. In a wide range of applications that spans chemistry \citep{Ning2014}, macroeconomics \citep{yamada2017estimating}, environmental science \citep{brantley2014mobile}, and medical sciences \citep{pettersson2013algorithm, marandi2015qualitative}, scalar functions of time $y(t)$ are observed and assumed to be a superposition of an underlying slowly varying baseline trend $\theta(t)$, other more rapidly varying components $s(t)$, and noise. In practice, $y(t)$ is observed at discrete time points $t_1, \ldots, t_n$, and we model the vector of samples $\VE{y}{i} = y(t_i)$ as
\begin{eqnarray*}
	\V{y} & = & \V{\theta} + \V{s} + \V{\varepsilon},
\end{eqnarray*}
where $\VE{\theta}{i} = \theta(t_i)$, $\VE{s}{i} = s(t_i)$, and $\V{\varepsilon} \in \Real^n$ is a vector of uncorrelated noise. For notational simplicity, for the rest of the paper, we assume that the time points take on the values $t_i = i$, but it is straightforward to generalize to an arbitrary grid of time points.

In some applications, the slowly varying component $\V{\theta}$ is the signal of interest, and the transient component $\V{s}$ is a vector of nuisance parameters. In our air quality application, the roles of $\V{\theta}$ and $\V{s}$ are reversed; $\V{s}$ represents the signal of interest and $\V{\theta}$ represents a baseline drift that obscures the identification of the important transient events encoded in $\V{s}$.

@@

In the rest of the paper, we detail our quantile trend filtering algorithms (Section \ref{sec:methods}) as well as how to choose the smoothing parameter (Section \ref{sec:lambda_choice}). We demonstrate through simulation studies that our proposed model provides better or comparable estimates of non-parametric quantile trends than existing methods (Section \ref{sec:simluation}). We further show that quantile trend filtering is a more effective method of drift removal for low-cost air quality sensors and results in improved signal classification compared to quantile smoothing splines (Section \ref{sec:application}). Finally, we discuss potential extensions of quantile trend filtering (Section \ref{sec:discussion}).

@@

In this application, as well as those described in \cite{Ning2014}, \cite{marandi2015qualitative}, and \cite{pettersson2013algorithm}, the goal is to estimate the trend in the baseline not the mean. We can define the trend in the baseline as the trend in a low quantile of the data. A variety of methods for estimating quantile trends have already been proposed. \cite{Koenker1978} were the first to propose substituting the sum-of-squares term with the check function  \Eqn{check} to estimate a conditional quantile instead of the conditional mean. Later, \cite{KoenkerNgPortnoy1994} proposed quantile trend filtering with $k = 1$ producing quantile trends that are piecewise linear, but they did not consider extensions to higher order differences. Rather than using the $\ell_1$-norm to penalize the discrete differences, \cite{nychka1995nonparametric} used the smoothing spline penalty based on the square of the $\ell_2$-norm:

\begin{equation*}
\label{eq:smoothingspline}
\sum_{i=1}^n\rho_{\tau}(y(t_i) - \theta(t_i)) + \lambda\int \theta''(t)^2 dt,
\end{equation*}
where $\theta(t)$ is a smooth function of time and $\lambda$ is a tuning parameter that controls the degree of smoothing. \cite{Oh2011} proposed an algorithm for solving the quantile smoothing spline problem by approximating the check function with a differentiable function. \cite{Racine2017} propose a method for estimating quantile trends that does employ the check function. In their method, the response is constrained to follow a location scale model and the conditional quantiles are estimated by combining Gaussian quantile functions with a kernel smoother and solving a local-linear least squares problem.

@@

We combine the ideas of quantile regression and trend filtering. For a single quantile level $\tau$, the quantile trend filtering problem is given in \Eqn{quantile_trend}. As with classic quantile regression, the quantile trend filtering problem is a linear program which can be solved by a number of methods. We want to estimate multiple quantiles simultaneously and to ensure that our quantile estimates are valid by enforcing the constraint that if $\tau_2 > \tau_1$ then $Q(\tau_2) \ge Q(\tau_1)$ where $Q$ is the quantile function of $\V{y}$. Even if a single quantile is ultimately desired, ensuring non-crossing allows information from nearby quantiles to be used to improve the estimates as we will see in the peak detection experiments in \Sec{peak_detection}. Given quantiles $\tau_1 < \tau_2 < \ldots < \tau_J$, the optimization problem becomes
\begin{eqnarray}
	\label{eq:noncross_trend}
	&&\underset{\M{\Theta} \in \mathcal{C}}{\text{minimize}}\; \sum_{j=1}^J \left [\rho_{\tau_j}(\V{y} - \V{\theta}_{j}) +
	\lambda_j \lVert \Mn{D}{k+1} \V{\theta}_j \rVert_1 \right ] 
\end{eqnarray}
where $\M{\Theta} = \begin{pmatrix} \V{\theta}_1 & \V{\theta}_2 & \cdots & \V{\theta}_J \end{pmatrix} \in \Real^{n \times J}$ is a matrix whose $j$th column corresponds to the $j$th quantile signal $\V{\theta}_j$ and the set $\mathcal{C} = \{\M{\Theta} \in \Real^{n \times J}: \ME{\theta}{ij} \leq \ME{\theta}{ij'} \;\text{for $j \leq j'$}\}$ encodes the non-crossing quantile constraints. The additional non-crossing constraints are linear inequalities involving the parameters, so the non-crossing quantile trends can still be estimated by a number of available linear programming solvers. We allow for the possibility that the degree of smoothness in the trends varies by quantile by allowing the smoothing parameter to vary with quantile as well. In the rest of this paper, we use $k=2$ to produce piecewise quadratic polynomials and report numerical results using the commercial solver Gurobi \citep{gurobi} and its R package implementation. However, we could easily substitute a free solver such as the Rglpk package by \cite{rglpk}.

@@

\Alg{admm} has the following convergence guarantees.
\begin{proposition}
	\label{prop:convergence}
	Let $\{\{\Mn{\Theta}{w}_m\}, \M{\Theta}_m\}$ denote the $m$th collection of iterates generated by \Alg{admm}. Then (i)
	$\lVert \Mn{\Theta}{w}_m - \Mn{U}{w}\M{\Theta}_m \rVert_{\text{F}} \rightarrow 0$ and (ii) $p_m \rightarrow p^\star$, where $p^\star$ is the optimal objective function value of
	\Eqn{quantile_windows} and $p_m$ is the objective function value of \Eqn{quantile_windows} evaluated at $\{\{\Mn{\Theta}{w}_m\}, \M{\Theta}_m\}$.
\end{proposition}
The proof of \Prop{convergence} is a straightforward application of the convergence result presented in Section 3.2 of \cite{boyd2011distributed}.

@@

\begin{eqnarray}
	r_{\text{primal}}^{m} & < &\epsilon_{\text{abs}}\sqrt{nJ} + \epsilon_{\text{rel}}\,\underset{w}{\max} \left[\max
	\left\{\lVert\Mn{\Theta}{w}_m \rVert_{\text{F}}, \lVert \M{\Theta}_{m} \rVert_{\text{F}} \right\}\right] \nonumber \\%\lVert \M{\Theta}_{k} \rVert_{\text{F}}} \right )\right]\\
	r_{\text{dual}}^{m} & < & \epsilon_{\text{abs}}\sqrt{nJ} + \epsilon_{\text{rel}}\,\sqrt{\sum_{w=1}^W
		\lVert \Mn{\Omega}{w}_m \rVert_{\text{F}}^2}.
\label{eq:stopping}
\end{eqnarray}

>>>>

% use option [preprint] to remove info line at bottom
% journal options: aop,aap,aos,aoas,ssy
% natbib option: authoryear
\documentclass[aoas]{imsart}
\RequirePackage{amsthm,amsmath,amsbsy,amsfonts}
\usepackage{bbm}
\usepackage{bm,graphicx,psfrag,epsf}
\usepackage{natbib}
\usepackage{url} % not crucial - just used below for the URL
\usepackage{placeins}
\usepackage{algorithm}
\usepackage{comment}
\usepackage{algpseudocode}
\usepackage{subfig} %% For \subfloat command
\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}
% provide arXiv number if available:
%\arxiv{arXiv:0000.0000}

@@

\section{Introduction}
\label{sec:intro}
In the last decade, low cost and portable air quality sensors have enjoyed dramatically increased usage. These sensors can provide an un-calibrated measure of a variety of pollutants in near real time, but deriving meaningful information from sensor data remains a challenge \citep{snyder2013changing}. For example, the ``SPod" is a low-cost sensor currently being investigated by researchers at the U.S. Environmental Protection Agency to detect volatile organic compound (VOC) emissions from industrial facilities \citep{thoma2016south}. Due to changes in temperature and relative humidity, the output signal exhibits a slowly varying baseline drift on the order of minutes to hours. Figure \ref{fig:raw_spod} provides an example of measurements from three SPod sensors co-located at the border of an industrial facility. All of the sensors respond to the pollutant signal, which is illustrated by the three sharp transient spikes at 11:32, 14:10, and 16:03. However, the baseline drift varies from one sensor to another, obscuring the detection of the peaks that alert the intrusion of pollutants. We show later that by estimating the baseline drift in each sensor and  removing it from the observed signals, peaks can be reliably detected from concordant residual signals from a collection of SPods using a simple data-driven thresholding strategy. Thus, accurately demixing a noisy observed time series into a slowly varying component and a transient component can lead to greatly improved and simplified downstream analysis.

@@

While this work is motivated by the analysis of data from low cost air quality sensors, the problem of demixing noisy time series into trends and transients is ubiquitous across many applications. \color{red} This includes electrocardiogram data \citep{luo2013hierarchical}, electrooculographic data (recording of eye movements) \citep{pettersson2013algorithm, marandi2015qualitative}, mass spectrometry \citep{Du2006,coombes2007pre}, chromatography \citep{Ning2014}, macroeconomics \citep{yamada2017estimating}, environmental science \citep{mecozzi2014polynomial,brantley2014mobile}, laser-induced breakdown spectroscopy \citep{franco2018comparison}, and astronomy \citep{Oh2004period}. In most instances, scalar functions of time $y(t)$ are observed and assumed to be a superposition of (i) an underlying slowly varying baseline trend $\theta(t)$, (ii) other more rapidly varying components $s(t)$, and (iii) noise. For example, in electrocardiogram data, one wishes to separate a train of transient waveforms $s(t)$ reflecting  the electrical activity of the heart from a nuisance ``baseline wander" $\theta(t)$ due a patient's respiration and other movements.\color{black}

In practice, $y(t)$ is observed at discrete time points $t_1, \ldots, t_n$, and we model the vector of samples $\VE{y}{i} = y(t_i)$ as
\begin{eqnarray*}
	\V{y} & = & \V{\theta} + \V{s} + \V{\varepsilon},
\end{eqnarray*}
where $\VE{\theta}{i} = \theta(t_i)$, $\VE{s}{i} = s(t_i)$, and $\V{\varepsilon} \in \Real^n$ is a vector of uncorrelated noise. For notational simplicity, for the rest of the paper, we assume that the time points take on the values $t_i = i$, but it is straightforward to generalize to an arbitrary grid of time points. In some applications, the slowly varying component $\V{\theta}$ is the signal of interest, and the transient component $\V{s}$ is a vector of nuisance parameters. In our air quality application, the roles of $\V{\theta}$ and $\V{s}$ are reversed; $\V{s}$ represents the signal of interest and $\V{\theta}$ represents a baseline drift that obscures the identification of the important transient events encoded in $\V{s}$.

@@

In the rest of the paper, we detail our quantile trend filtering algorithms (Section \ref{sec:methods}) and investigate how to choose the smoothing parameter (Section \ref{sec:lambda_choice}). We demonstrate through simulation studies that our proposed methods provide better or comparable estimates of non-parametric quantile trends than existing methods (Section \ref{sec:simluation}). We further show that quantile trend filtering is a more effective method of drift removal for low-cost air quality sensors and results in improved signal classification compared to quantile smoothing splines (Section \ref{sec:application}). Finally, we discuss potential extensions of quantile trend filtering (Section \ref{sec:discussion}).

@@

In this application, as well as those described in \cite{Ning2014}, \cite{marandi2015qualitative}, and \cite{pettersson2013algorithm}, the goal is to estimate the trend in the baseline, not the mean. We can define the trend in the baseline as the trend in a low quantile of the data. A variety of methods for estimating quantile trends have already been proposed. \cite{Koenker1978} were the first to propose substituting the sum-of-squares term with the check function  \Eqn{check} to estimate a conditional quantile instead of the conditional mean. Later, \cite{KoenkerNgPortnoy1994} proposed quantile trend filtering with $k = 1$ producing quantile trends that are piecewise linear, but they did not consider extensions to higher order differences. Rather than using the $\ell_1$-norm to penalize the discrete differences, \cite{nychka1995nonparametric} used the smoothing spline penalty based on the square of the $\ell_2$-norm:

\begin{equation*}
\label{eq:smoothingspline}
\sum_{i=1}^n\rho_{\tau}(y(t_i) - \theta(t_i)) + \lambda\int \theta''(t)^2 dt,
\end{equation*}
where $\theta(t)$ is a smooth function of time, and $\lambda$ is a tuning parameter that controls the degree of smoothing. \cite{Oh2011} proposed an algorithm for solving the quantile smoothing spline problem by approximating the check function with a differentiable function. \cite{Racine2017} propose a method for estimating quantile trends that does not employ the check function. In their method, the response is constrained to follow a location scale model and the conditional quantiles are estimated by combining Gaussian quantile functions with a kernel smoother and solving a local-linear least squares problem.

@@

We combine the ideas of quantile regression and trend filtering. For a single quantile level $\tau$, the quantile trend filtering problem is given in \Eqn{quantile_trend}. As with classic quantile regression, the quantile trend filtering problem is a linear program which can be solved by a number of methods. We want to estimate multiple quantiles simultaneously and to ensure that our quantile estimates are valid by enforcing the constraint that if $\tau_2 > \tau_1$ then $Q(\tau_2) \ge Q(\tau_1)$, where $Q$ is the quantile function of $\V{y}$. Even if a single quantile is ultimately desired, ensuring non-crossing allows information from nearby quantiles to be used to improve the estimates, as we will see in the peak detection experiments in \Sec{peak_detection}. Given quantiles $\tau_1 < \tau_2 < \ldots < \tau_J$, the optimization problem becomes
\begin{eqnarray}
	\label{eq:noncross_trend}
	&&\underset{\M{\Theta} \in \mathcal{C}}{\text{minimize}}\; \sum_{j=1}^J \left [\rho_{\tau_j}(\V{y} - \V{\theta}_{j}) +
	\lambda_j \lVert \Mn{D}{k+1} \V{\theta}_j \rVert_1 \right ] 
\end{eqnarray}
where $\M{\Theta} = \begin{pmatrix} \V{\theta}_1 & \V{\theta}_2 & \cdots & \V{\theta}_J \end{pmatrix} \in \Real^{n \times J}$ is a matrix whose $j$th column corresponds to the $j$th quantile signal $\V{\theta}_j$ and the set $\mathcal{C} = \{\M{\Theta} \in \Real^{n \times J}: \ME{\theta}{ij} \leq \ME{\theta}{ij'} \;\text{for $j \leq j'$}\}$ encodes the non-crossing quantile constraints. The additional non-crossing constraints are linear inequalities involving the parameters, so the non-crossing quantile trends can still be estimated by a number of available linear programming solvers. We allow for the possibility that the degree of smoothness in the trends varies by quantile by allowing the smoothing parameter to vary with quantile. In the rest of this paper, we use $k=2$ to produce piecewise quadratic polynomials and report numerical results using the commercial solver Gurobi \citep{gurobi} and its R package implementation. However, we could easily substitute a free solver such as the Rglpk package by \cite{rglpk}.

@@

\begin{comment}
\Alg{admm} has the following convergence guarantees.
\begin{proposition}
	\label{prop:convergence}
	Let $\{\{\Mn{\Theta}{w}_m\}, \M{\Theta}_m\}$ denote the $m$th collection of iterates generated by \Alg{admm}. Then (i)
	$\lVert \Mn{\Theta}{w}_m - \Mn{U}{w}\M{\Theta}_m \rVert_{\text{F}} \rightarrow 0$ and (ii) $p_m \rightarrow p^\star$, where $p^\star$ is the optimal objective function value of
	\Eqn{quantile_windows} and $p_m$ is the objective function value of \Eqn{quantile_windows} evaluated at $\{\{\Mn{\Theta}{w}_m\}, \M{\Theta}_m\}$.
\end{proposition}
The proof of \Prop{convergence} is a straightforward application of the convergence result presented in Section 3.2 of \cite{boyd2011distributed}.

@@

\begin{eqnarray}
	r_{\text{primal}}^{m} & < &\epsilon_{\text{abs}}\sqrt{nJ} + \epsilon_{\text{rel}}\,\underset{w}{\max} \left[\max
	\left\{\lVert\Mn{\Theta}{w}_m \rVert_{\text{F}}, \lVert \M{\Theta}_{m} \rVert_{\text{F}} \right\}\right] \nonumber \\%\lVert \M{\Theta}_{k} \rVert_{\text{F}}} \right )\right]\\
	r_{\text{dual}}^{m} & < & \epsilon_{\text{abs}}\sqrt{nJ} + \epsilon_{\text{rel}}\,\sqrt{\sum_{w=1}^W
		\lVert \Mn{\Omega}{w}_m \rVert_{\text{F}}^2}.
\label{eq:stopping}
\end{eqnarray}
\end{comment}

The windowed quantile trend filtering optimization problem \Eqn{quantile_windows} opens the door to significant computational gains over solving the original quantile trend filtering problem \Eqn{noncross_trend}. The quantile trend filtering problem for a single window is a linear program with $N\times J$ parameters (number of observations by number of quantiles), which can be solved in computational time proportional to $(NJ)^3$.  Consequently, solving a large problem using \Alg{admm} should require less computational time than solving \Eqn{noncross_trend}, even if the sub-problems are solved sequentially. We demonstrate the advantages of \Alg{admm} through timing experiments (\Fig{timing}). For each data size, 25 datasets were simulated using the peaks simulation design described below. Trends for the fifth, tenth, and fifteenth quantiles were fit simultaneously using $\lambda_j = n/5$ for all $j$. We use from one to four windows for each data size with an overlap of 500. %\Alg{admm} was terminated using a standard stopping rule for ADMM, which is described in the Supplementary Materials. %when \Eqn{stopping} was satisfied, defining $\epsilon_{abs} = 0.01$ and $\epsilon_{rel} = 0.001$. 
\Fig{timing} shows that using 4 windows instead of one on data sizes of 55,000 provides a factor of 3 decrease in computation time. The timing experiments were conducted on an Intel Xeon based Linux cluster using two processor cores.
